---
title: 'Last words of Texas death row inmates'
subtitle: 'An exploratory text analysis'
author: 'Edward Yu'
date: '2019-03-23'
output:
  html_document:
      theme: flatly
      highlight: tango
      toc: true
      toc_float:
        collapsed: true
      toc_depth: 3
      df_print: paged
      code_folding: show
      # fig_width: 7
      # fig_height: 6
      # fig_caption: true
      # citation_package: natbib
# bibliography: bib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo    = TRUE,
                      error   = FALSE,
                      message = FALSE,
                      warning = FALSE)

if(!require(tidyverse)) install.packages("tidyverse")
library(tidyverse)
if(!require(naniar)) install.packages("naniar")
library(naniar) # gg_miss
if(!require(ggridges)) install.packages("ggridges")
library(ggridges)
if(!require(viridis)) install.packages("viridis")
library(viridis)
if(!require(gridExtra)) install.packages("gridExtra")
library(gridExtra)

# clear history
rm(list = ls())

# setwd("I:/Code-CAD/R/bakery.sales")
setwd("D:/Code/R/inmate.last.words")
```

# Introduction
I found this data via a Kaggle post: https://www.kaggle.com/mykhe1097/last-words-of-death-row-inmates. The original data was taken from: https://www.tdcj.texas.gov/death_row/dr_executed_offenders.html.  

The title alone merits more than a passing glance. Moving forward I'll start with a very simple, generalized analysis, moving on to a more in depth sentiment analysis utilizing the tidytext package.

# Import
In general my beginning process is very similar from project to project: we import the data, `glimpse` it to get an idea of the structure, sort all variables by descending unique levels, and plot the missingness of each variables.    



```{r}
# import
x <- read_csv("data/Texas Last Statement - CSV.csv")

glimpse(x)

# Find unique levels
x.levels <- cbind(colnames(x),
                  (as.data.frame(sapply(x,function(x) length(unique(x))))))
colnames(x.levels) <- c("var","levels")
row.names(x.levels) <- NULL
x.levels[order(-x.levels[,2]),]

# % missingness using naniar package
gg_miss_var(x, show_pct = T)
```

